{\Huge Supersymmetry} \\ \vspace{5mm} Lecture notes for FYS5190/FYS9190 Supersymmetry

### 2025-11-13 mssm.tex
Changes for lecture 13.11.25

%%%
\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{feynman}
%
\diagram [layered layout, horizontal=a to b] {
a [particle=$B$] -- [boson] b,
b -- [boson] c [particle=$B$],
b -- [boson] d [particle=$B$],
};
%
\diagram [xshift=6cm, small, horizontal=a to t1] {
a [particle=$B$] -- [boson] t1 -- [fermion] t2 -- [fermion, edge label=$f$] t3 -- [fermion] t1,
t2 -- [boson] p1 [particle=$B$],
t3 -- [boson] p2 [particle=$B$],
p1 -- [opacity=0] p2,
};
\end{feynman}
\end{tikzpicture}
\caption{The tree level coupling between three gauge bosons $B$ (left), and the one-loop fermion contribution to the same process (right).}  
\label{fig:anomaly}
\end{figure}
%%%

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Renormalisation group equations}
\label{sec:RGE}
%%%%%%%%%%%%%%%%%%%%%%
Renormalisation, the removal of infinities from field theory predictions, introduces a fixed scale $\mu$ at which the fields and the parameters of the Lagrangian, the couplings, are defined. For example, the charge of the electron is not simply the {\bf bare charge} $e_0$ given in the original Lagrangian, but a charge at a given energy scale $\mu$, $e(\mu)$, which is the scale at which the theory wants to describe the electron, and which we can measure in an experiment at that scale. Describing the scattering an electron at very high energy will require a different value of $e(\mu)$ than at a low energy. This scale dependence in the coupling is an experimentally well verified fact.\footnote{It is also impossible to avoid if we accept that the electron is a point particle. Since the potential has the form $V(r)\propto e/r$ an infinite energy would appear unless we were somehow to modify the charge at high energies, or equivalently, short distances.} The relationship between a bare field or coupling, and the (dressed) renormalised field or coupling can be found from the so-called {\bf renormalisation constant} $Z$ that renormalises the parameter. For example for a field $\phi$, $\phi=Z_\phi \phi_0$, and for coupling $g$, $g=Z_gg_0$.

However, since $\mu$ is not an observable {\it per se} but in principle a choice of how to write down the theory to compare to an experiment (at which energy scale to write down the Lagrangian), the renormalised effective action $S$ for a physical process should be invariant under a change of $\mu$, which is expressed as:\footnote{It is more common in the literature to find this expressed in terms of the Green's function $G^{(n)}$ for a given $n$-point correlation, {\it i.e.}\ a process with $n$-field insertions. In this form the equation is known as the {\bf Callan--Symanzik equation}.}
\begin{equation}
\mu \frac{d}{d\mu}S(\Phi, \lambda, \mu) = 0,
\end{equation}
where $\lambda$ is a generic name for the couplings of the theory and $\Phi$ represents the (super)fields that have been renormalised.\footnote{In Sec.~\ref{sec:vacuum_energy} we mentioned how the non-renormalisation theorem implies that we do not need to renormalise the coupling constants of the superpotential separately. In the MSSM this is the $\mu$ coupling (not to be confused with the energy scale here also called $\mu$) and the Yukawa couplings. This will now have the consequence that their renormalisation can be expressed in terms of the renormalisation of the fields.} This equation can be re-written in terms of partial derivatives
\begin{equation}
\left(\mu \frac{\partial}{\partial \mu} + \beta_\lambda\frac{\partial}{\partial\lambda}+n_\Phi\gamma_\Phi\right)S(\Phi, \lambda, \mu)=0,
\end{equation}
which is the {\bf renormalisation group equation} (RGE). Here $\beta_\lambda$ is the {\bf $\boldsymbol \beta$-function}:
\begin{equation}
\beta_\lambda\equiv\mu\frac{\partial \lambda}{\partial \mu}.
\end{equation}
which describes the behaviour of a Lagrangian parameter $\lambda$ as a function of the energy scale $\mu$ away from the value where it was defined, often denoted $\mu_0$. The {\bf anomalous dimension} $\boldsymbol\gamma$ describes the scaling of the fields and the factor $n$ gives the number of fields (of a given kind) in the effective action. The RGE balances the different contributions to sum to zero.

@@

Something funny happens to the gaugino soft-mass parameters $M_i$ if we look at their running. From (\ref{eq:Mi_beta}) the one-loop $\beta$-functions for the $M_i$ in the MSSM are
\begin{equation}
\beta_{M_i}|_{\rm 1-loop} \equiv \frac{d}{dt}M_i = \frac{1}{8\pi^2}g_i^2M_ib_i.
\end{equation}
As a consequence all three ratios $M_i/g_i^2$ are scale independent at one loop. To see this let $R=M_i/g_i^2$, then
\begin{equation}
\beta_R \equiv \frac{dR}{dt}= \frac{\frac{dM_i }{dt}g_i^2 - M_i \frac{d}{dt} g_i^2}{g_i^4} = \frac{\frac{1}{8\pi^2}g_i^2M_ib_i\cdot g_i^2 - M_i\cdot 2g_i\cdot\frac{1}{16\pi} g_i^3 b_i}{g_i^4} = 0.
\end{equation}
In other words, $R$ does not change with scale $t$.

If we now use that the coupling constants unify at the GUT scale to the coupling $g_u$, and assume that the soft-masses are the same at that scale $m_{1/2} = M_1(\mu_\text{GUT}) = M_2(\mu_\text{GUT}) = M_3(\mu_\text{GUT})$,\footnote{Again, not unreasonable if the spontaneous symmetry breaking mechanism acts uniformly for all the gauginos.} it follows that
\begin{equation}
\frac{M_1}{g_1^2} = \frac{M_2}{g_2^2} = \frac{M_3}{g_3^2} = \frac{m_{1/2}}{g_u^2},
\end{equation}
at all scales!\footnote{At one-loop level.} This is a very powerful and predictive assumption. Because of the relationship between the electroweak couplings and the electric charge, $e=g'\cos\theta_W=g\sin\theta_W$, it leads to the following relation
\begin{equation}
M_3 = \frac{\alpha_s}{\alpha}\sin^2\theta_WM_2 = \frac{3}{5}\frac{\alpha_s}{\alpha}\cos^2\theta_W M_1,
\end{equation}
which, inserting values for the fine structure constant, the strong coupling, and the Weinberg angle, numerically predicts
\[M_3:M_2:M_1 \simeq 6:2:1\]
at the electroweak scale. We will return to the implications of this when discussing the gauginos in Sec.~\ref{sec:electroweakinos}.

In Fig.~\ref{fig:MSSMrun}, again taken from Martin~\cite{Martin:1997ns}, we show the running of all the types of soft parameters in the MSSM. We assume unified soft-mass parameters $m_{1/2}$ for the gauginos and $m_0$ for the Higgs and sfermions at the GUT scale. Shown is the gaugino mass parameters $M_i$ (solid black), the Higgs mass parameters $m_{H_{d/u}}^2$ (dot-dashed green), the third generation sfermion soft terms $m_{d_3}$, $m_{Q_3}$, $m_{u_3}$, $m_{L_3}$ and $m_{e_3}$ (dashed red and blue, listed from top to bottom), and the corresponding first and second generation terms (solid red and blue lines).
%%%%%%%%% 
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{figures/MSSMrun.eps} 
\caption{The RGE evolution of soft-mass parameters in the MSSM with typical minimal supergravity-inspired boundary conditions imposed at $2\cdot10^{16}$\,GeV. The parameter values
used for this illustration were $m_0 = 200$\,GeV, $m_{1/2 }= -A_0 =600$\,GeV, $\tan\beta = 10$, and ${\rm sgn}(\mu)=+$.}
\label{fig:MSSMrun}
\end{figure}
%%%%%%%%%%

>>>>

%%%
\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{feynman}
%
\diagram [layered layout, horizontal=a to b] {
a [particle=$A$] -- [boson] b,
b -- [boson] c [particle=$V$],
b -- [boson] d [particle=$V$],
};
%
\diagram [xshift=6cm, small, horizontal=a to t1] {
a [particle=$A$] -- [boson] t1 -- [fermion] t2 -- [fermion, edge label=$f$] t3 -- [fermion] t1,
t2 -- [boson] p1 [particle=$V$],
t3 -- [boson] p2 [particle=$V$],
p1 -- [opacity=0] p2,
};
\end{feynman}
\end{tikzpicture}
\caption{The tree level coupling between three gauge bosons with vector $V$ and axial vector $A$ couplings (left), and the one-loop fermion contribution to the same process (right).}  
\label{fig:anomaly}
\end{figure}
%%%

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Renormalisation group equations}
\label{sec:RGE}
%%%%%%%%%%%%%%%%%%%%%%
Renormalisation, the removal of infinities from field theory predictions, introduces a fixed scale $\mu$ at which the fields and the parameters of the Lagrangian, the couplings, are defined. For example, the charge of the electron is not simply the {\bf bare charge} $e_0$ given in the original Lagrangian, but a charge at a given energy scale $\mu$, $e(\mu)$, which is the scale at which the theory wants to describe the electron, and which we can measure in an experiment at that scale. Describing the scattering an electron at very high energy will require a different value of $e(\mu)$ than at a low energy. This scale dependence in the coupling is an experimentally well verified fact.\footnote{It is also impossible to avoid if we accept that the electron is a point particle. Since the Coulomb potential has the form $V(r)\propto e/r$, an infinite energy would appear unless we were somehow to modify the charge at high energies, or equivalently, short distances.} The relationship between a bare field or coupling, and the (dressed) renormalised field or coupling can be found from the so-called {\bf renormalisation constant} $Z$ that renormalises the parameter. For example for a field $\phi$, $\phi=Z_\phi \phi_0$, and for coupling $g$, $g=Z_gg_0$.

However, since $\mu$ is not an observable {\it per se} but in principle a choice of how to write down the theory to compare to an experiment (at which energy scale to write down the Lagrangian), the renormalised effective action $S$ for a physical process should be invariant under a change of $\mu$, which is expressed as:\footnote{It is more common in the literature to find this expressed in terms of the Green's function $G^{(n)}$ for a given $n$-point correlation, {\it i.e.}\ a process with $n$-field insertions. In this form the equation is known as the {\bf Callan--Symanzik equation}.}
\begin{equation}
\mu \frac{d}{d\mu}S(\Phi, \lambda, \mu) = 0,
\end{equation}
where $\lambda$ is a generic name for the couplings of the theory and $\Phi$ represents the (super)fields that have been renormalised.\footnote{In Sec.~\ref{sec:vacuum_energy} we mentioned how the non-renormalisation theorem implies that we do not need to renormalise the coupling constants of the superpotential separately. In the MSSM this is the $\mu$ coupling (not to be confused with the energy scale here also called $\mu$) and the Yukawa couplings. This will now have the consequence that their renormalisation can be expressed in terms of the renormalisation of the fields.}

Starting from the relationship between the renormalised action $S_R$ and the bare action $S_0$,
\[ S_R(\Phi,\lambda,\mu)= Z_\Phi^{n_\Phi/2} S_0(\Phi_0,\lambda_0,\mu),\]
this equation can be re-written in terms of partial derivatives
\begin{equation}
\left(\mu \frac{\partial}{\partial \mu} + \beta_\lambda\frac{\partial}{\partial\lambda}+n_\Phi\gamma_\Phi\right)S(\Phi, \lambda, \mu)=0,
\end{equation}
which is the {\bf renormalisation group equation} (RGE). Here $\beta_\lambda$ is the {\bf $\boldsymbol \beta$-function}:
\begin{equation}
\beta_\lambda\equiv\mu\frac{\partial \lambda}{\partial \mu}.
\end{equation}
which describes the behaviour of a Lagrangian parameter $\lambda$ as a function of the energy scale $\mu$ away from the value where it was defined, often denoted $\mu_0$. The {\bf anomalous dimension} of the field
\begin{equation}
\boldsymbol\gamma =  \mu\frac{\partial}{\partial\mu}\ln Z^\half_\Phi,
\end{equation} 
describes the scaling of the fields and the factor $n_\Phi$ gives the number of external fields (of a given kind) in the effective action. The RGE balances the different contributions to sum to zero.

@@

Something funny happens to the gaugino soft-mass parameters $M_i$ if we look at their running. From (\ref{eq:Mi_beta}) the one-loop $\beta$-functions for the $M_i$ in the MSSM are
\begin{equation}
\beta_{M_i}|_{\rm 1-loop} \equiv \frac{d}{dt}M_i = \frac{1}{8\pi^2}g_i^2M_ib_i.
\end{equation}
As a consequence all three ratios $M_i/g_i^2$ are scale independent at one loop. To see this let $R_i=M_i/g_i^2$, then
\begin{equation}
\beta_{R_i} \equiv \frac{dR_i}{dt}= \frac{\frac{dM_i }{dt}g_i^2 - M_i \frac{d}{dt} g_i^2}{g_i^4} = \frac{\frac{1}{8\pi^2}g_i^2M_ib_i\cdot g_i^2 - M_i\cdot 2g_i\cdot\frac{1}{16\pi} g_i^3 b_i}{g_i^4} = 0.
\end{equation}
In other words, $R_i$ does not change with scale $t$.

If we now use that the coupling constants unify at the GUT scale to the coupling $g_u$, and assume that the soft-masses are the same at that scale $m_{1/2} = M_1(\mu_\text{GUT}) = M_2(\mu_\text{GUT}) = M_3(\mu_\text{GUT})$,\footnote{Again, not unreasonable if the spontaneous symmetry breaking mechanism acts uniformly for all the gauginos.} it follows that
\begin{equation}
\frac{M_1}{g_1^2} = \frac{M_2}{g_2^2} = \frac{M_3}{g_3^2} = \frac{m_{1/2}}{g_u^2},
\end{equation}
at all scales!\footnote{At one-loop level.} This is a very powerful and predictive assumption. Because of the relationship between the electroweak couplings and the electric charge, $e=g'\cos\theta_W=g\sin\theta_W$, it leads to the following relation
\begin{equation}
M_3 = \frac{\alpha_s}{\alpha}\sin^2\theta_WM_2 = \frac{3}{5}\frac{\alpha_s}{\alpha}\cos^2\theta_W M_1,
\end{equation}
which, inserting values for the fine structure constant $\alpha$, the strong coupling $\alpha_s$, and the Weinberg angle $\sin^2\theta_W$, numerically predicts
\[M_3:M_2:M_1 \simeq 6:2:1\]
at the electroweak scale. We will return to the implications of this when discussing the gauginos in Sec.~\ref{sec:electroweakinos}.

In Fig.~\ref{fig:MSSMrun}, again taken from Martin~\cite{Martin:1997ns}, we show the running of all the types of soft parameters in the MSSM. We assume unified soft-mass parameters $m_{1/2}$ for the gauginos and $m_0$ for the Higgs and sfermions at the GUT scale. Shown is the gaugino mass parameters $M_i$ (solid black), the Higgs mass parameters $m_{H_{d/u}}^2$ (dot-dashed green), the third generation sfermion soft terms $m_{d_3}$, $m_{Q_3}$, $m_{u_3}$, $m_{L_3}$ and $m_{e_3}$ (dashed blue and red, listed from top to bottom), and the corresponding first and second generation terms $m_{d_1}=m_{d_2}$, $m_{Q_1}=m_{Q_2}$, $m_{u_1}=m_{u_2}$, $m_{L_1}=m_{L_2}$ and $m_{e_1}=m_{e_2}$ (solid blue and red lines).
%%%%%%%%% 
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{figures/MSSMrun.eps} 
\caption{The RGE evolution of soft-mass parameters in the MSSM with typical minimal supergravity-inspired boundary conditions imposed at $2\cdot10^{16}$\,GeV. The parameter values
used for this illustration were $m_0 = 200$\,GeV, $m_{1/2 }= -A_0 =600$\,GeV, $\tan\beta = 10$, and ${\rm sgn}(\mu)=+$.}
\label{fig:MSSMrun}
\end{figure}
%%%%%%%%%%
